---
title: "BeachCOMBERS for NMS infographics"
output: 
  html_document: 
    toc: yes
    toc_float: true
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = F, warning = F)
```

## Data


* `bm16`: [b]irds + [m]ammals, up to 20[16]
  - [ERDDAP - BeachCOMBERS Effort-Based Marine Mammal and Seabird Beach Cast Survey - Data Access Form](https://oceanview.pfeg.noaa.gov/erddap/tabledap/EB_MM_BC.html) `EB_MM_BC`, birds **+ mammals**, but only up to **2016**
* `b019`: [b]irds [o]nly, up to 20[19]
  - [ERDDAP - Seabird beached mortality - Data Access Form](https://oceanview.pfeg.noaa.gov/erddap/tabledap/cciea_B_B_MORT.html) `cciea_B_B_MORT`
  - [researchworkspace.com/.../BeachCOMBERS_formatted_2019.csv](https://researchworkspace.com/file/5851070/BeachCOMBERS_formatted_2019.csv), **birds only**, up to **2019**
  - TODO: check ERDDAP same as researchworkspace.com

```{r}
# libraries
library(here)
library(glue)
library(stringr)
library(readr)
library(dplyr)
library(purrr)
library(DT)
library(rerddap)
library(sf)
library(mapview)
library(leafsync)
library(leaflet)
library(skimr)
library(lubridate)
library(plotly)
here <- here::here

# paths & variables
erddap_url <- "https://oceanview.pfeg.noaa.gov/erddap/"
d_id       <- "EB_MM_BC"
dir_data   <- here("data")
# manual download from Axiom space:
axiom_url  <- "https://researchworkspace.com/file/5851070/BeachCOMBERS_formatted_2019.csv"
# TODO: compsare BeachCOMBERS_formatted_2019.csv w/ ERDDAP cciea_B_B_MORT
d_bo19_csv <- glue("{dir_data}/{d_id}_bo19.csv")
d_bm16_csv <- glue("{dir_data}/{d_id}_bm16.csv")
d_csv      <- glue("{dir_data}/{d_id}.csv")
segs_geo   <- glue("{dir_data}/{d_id}_segments.geojson")
d_spp_csv  <- glue("{dir_data}/{d_id}_species.csv")
cols_bo19_csv       <- glue("{dir_data}/cols_bo19.csv")
cols_bm16tobo19_csv <- glue("{dir_data}/cols_bm16tobo19.csv")
d_mbnms_csv <- glue("{dir_data}/BeachCOMBERS_MBNMS.csv")

Sys.setenv(RERDDAP_DEFAULT_URL = erddap_url)

if (!file.exists(d_bo19_csv)){
  download.file(axiom_url, d_bo19_csv)
}

if (!file.exists(d_bm16_csv)){
  d_info <- info(d_id)
  d      <- tabledap(d_info)
  write_csv(d, d_bm16_csv)
}

d_bo19 <- read_csv(d_bo19_csv, guess_max = 100000)
d_bm16 <- read_csv(d_bm16_csv, guess_max = 10000)

if (!file.exists(cols_bm16tobo19_csv)){
  tibble(
    bo19_name = names(d_bo19),
    bo19_type = map_chr(d_bo19, ~ class(.) %>% paste(collapse = ","))) %>%
    arrange(bo19_name) %>%
    write_csv(cols_bo19_csv)
  tibble(
    bm16_name = names(d_bm16),
    bm16_type = map_chr(d_bm16, ~ class(.) %>% paste(collapse = ","))) %>%
    arrange(bm16_name) %>%
    write_csv(cols_bm16tobo19_csv)
  stop("NEXT: manually match column name and type for bm16 ({basename(d_bm16tobo19_csv)}) to match bo19 ({basename(d_bo19_csv)}).")
}
cx      <- read_csv(cols_bm16tobo19_csv)
cx_vars <- setNames(cx$bm16_name, cx$bo19_name)

# TODO: update bm16 rows
# table(d_bo19 %>% select(age_class_label, age_class))
# table(d_bo19 %>% select(condition_label, condition))
# table(d_bo19 %>% select(sex_label, sex))
# other bo19 flds missing in bo19: 
# - organization_code
# - organization_label
# - deposition_rate_carcass_per_km
# - inst_code

# setdiff(unique(d_bm16$age_class), unique(d_bo19$age_class))
# unique(d_bo19$age_class_label)

d_bm164bo19 <- d_bm16 %>% 
    filter(bird_or_mammal != "bird") %>% 
    mutate(
      unique_carcass_identifier = as.character(unique_carcass_identifier)) %>% 
    rename(!!cx_vars)
d <- bind_rows(
  d_bo19,
  d_bm164bo19) %>% 
  mutate(
    species = glue(
      "{species_code}: {species_common_name} ({species_sci_name}; ITIS: {species_itis})"))
```

Of the `r format(nrow(d), big.mark=",") ` rows in this dataset, let's look at the first 1,000 records:

```{r}
d %>% 
  slice(1:100) %>% 
  datatable()
```

```{r}
skimr::skim(d)
```

- `percent_of_beach_surveyed`: all NaN
- `cause_of_death`:
  >    0     1     2     3     4     5     6     7 
  >  818     6   302    21 64182   189     3     8

## taxa

```{r}
if (!file.exists(d_spp_csv)){
  
  flds_spp <- names(d) %>% 
    str_subset("species") %>% 
    c("bird_or_mammal")
  
  d_spp <- d %>% 
    select(!!!flds_spp) %>% 
    group_by_all() %>% 
    summarise(n = n()) %>% 
    arrange(bird_or_mammal, species)
  
  write_csv(d_spp, d_spp_csv)
}
d_spp <- read_csv(d_spp_csv)

table(d_spp$bird_or_mammal)
```

### birds

```{r}
d_spp %>% 
  filter(bird_or_mammal == "bird") %>% 
  datatable()
```

### mammals

```{r}
d_spp %>% 
  filter(bird_or_mammal == "mammal") %>% 
  datatable()
```

## spatial

### intersect sanctuary with segments

```{r}
xy2ln <- function(x1, y1, x2, y2){
  if (any(is.na(c(x1, y1, x2, y2)))) return(NA)
  
  st_linestring(c(
    st_point(c(x1, y1)), 
    st_point(c(x2, y2))))
}

if (!file.exists(segs_geo)){
  
  segs <- d %>% 
    filter(
      !is.na(north_longitude), !is.na(north_latitude),
      !is.na(south_longitude), !is.na(south_latitude)) %>% 
    group_by(
      beach_segment_code, 
      #longitude, latitude, 
      north_longitude, north_latitude,
      south_longitude, south_latitude) %>% 
    summarize(
      n_rows = n()) %>% 
    ungroup() %>% 
    mutate(
      geom = pmap(list(
        north_longitude, north_latitude,
        south_longitude, south_latitude), xy2ln) %>% 
        st_sfc(crs = 4326)) %>% 
    st_set_geometry(.$geom)
  
  write_sf(segs, segs_geo, delete_dsn = T)
}
segs <- read_sf(segs_geo)

# TODO: move used fxns into library and utility.R
source("~/github/cinms/scripts/rocky.R")

nms     <- "mbnms"
nms_ply <- get_nms_ply(nms)

nms_segs <- segs %>% 
  filter(
    st_intersects(
      segs, nms_ply, sparse = F))

m1 <- mapview(nms_ply) + 
  mapview(segs, color="red") + 
  mapview(nms_segs, color="green")
m1
```

NOTE the red segments near San Jose that did not get intersected by the polygon because of spatial coarseness and mismatch.

### intersect buffered sanctuary with segments

Buffer sanctuary by `0.005` decimal degress (~ `r 111 * 0.005` km at equator).

```{r}
nms_ply_buf <- st_buffer(nms_ply, dist = 0.005)

nms_buf_segs <- segs %>% 
  filter(
    st_intersects(
      segs, nms_ply_buf, sparse = F))

d_nms_buf <- d %>% 
  semi_join(
    nms_buf_segs, by = "beach_segment_code")

# write to csv
d_mbnms <- d_nms_buf %>% 
  mutate(
    year = year(time)) %>% 
  group_by(
    year,
    carcass_present, bird_or_mammal, 
    species_code, species_sci_name, species_common_name, species_itis, species, 
    age_class, sex, condition, cause_of_death, alive) %>% 
  summarize(
    n = n()) %>% 
  ungroup()
write_csv(d_mbnms, d_mbnms_csv)

d_mbnms <- read_csv(d_mbnms_csv)

m2 <- mapview(nms_ply_buf) +
  mapview(nms_buf_segs, color="green")

leafsync::sync
sync(m1, m2)
```

Filter by `beach_segment_code` in MBNMS: `r format(nrow(d_nms_buf), big.mark=",")` of `r format(nrow(d), big.mark=",")` rows (`r round(nrow(d_nms_buf)/nrow(d) * 100, 1) `%).

## plots

### bird carcasses

#### static

```{r}
g_birds <- d_mbnms %>% 
  filter(
    bird_or_mammal == "bird",
    carcass_present == T) %>% 
  group_by(year) %>% 
  summarize(
    n = sum(n)) %>% 
  ungroup() %>% 
  ggplot(aes(x=year, y=n)) + 
  #geom_bar(stat = "identity") #+ 
  geom_col() #+ 
  #title("Bird carcasses over time")
print(g_birds)
```

#### interactive, color by species on hover


```{r}
#plot.new()
g_bird_spp <- d_mbnms %>% 
  filter(
    bird_or_mammal == "bird",
    carcass_present == T) %>% 
  group_by(year, species) %>% 
  summarize(
    n = sum(n)) %>% 
  ungroup() %>% 
  ggplot(aes(x=year, y=n, fill=species)) + 
  #geom_bar(stat = "identity") + 
  geom_col() +
  #title("Bird carcasses over time, by species") + 
  theme(legend.position = "none")

ggplotly(g_bird_spp)
```

### mammal carcasses

#### static

```{r}
g_mammal <- d_mbnms %>% 
  filter(
    bird_or_mammal == "mammal",
    carcass_present == T) %>% 
  group_by(year) %>% 
  summarize(
    n = sum(n)) %>% 
  ungroup() %>%  
  ggplot(aes(x=year, y=n)) + 
  #geom_bar(stat = "identity") #+ 
  geom_col() #+ 
  #title("Bird carcasses over time")
print(g_mammal)
```

#### interactive, color by species on hover


```{r}
#plot.new()
g_mammal_spp <- d_mbnms %>% 
  filter(
    bird_or_mammal == "mammal",
    carcass_present == T) %>% 
 group_by(year, species) %>% 
  summarize(
    n = sum(n)) %>% 
  ungroup() %>% 
  ggplot(aes(x=year, y=n, fill=species)) + 
  #geom_bar(stat = "identity") + 
  geom_col() +
  #title("Bird carcasses over time, by species") + 
  theme(legend.position = "none")

ggplotly(g_mammal_spp)
```


