---
title: "BeachCOMBERS for NMS infographics"
output: 
  html_document: 
    toc: yes
    toc_float: true
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = F, warning = F)
```

## ERDDAP

* `EB_MM_BC`: [ERDDAP - BeachCOMBERS Effort-Based Marine Mammal and Seabird Beach Cast Survey - Data Access Form](https://oceanview.pfeg.noaa.gov/erddap/tabledap/EB_MM_BC.html)
* `cciea_B_B_MORT`: [ERDDAP - Seabird beached mortality - Data Access Form](https://oceanview.pfeg.noaa.gov/erddap/tabledap/cciea_B_B_MORT.html): subset of `EB_MM_BC`?

```{r}
# libraries
library(here)
library(glue)
library(stringr)
library(readr)
library(dplyr)
library(purrr)
library(DT)
library(rerddap)
library(sf)
library(mapview)
library(leaflet)

# paths & variables
erddap_url <- "https://oceanview.pfeg.noaa.gov/erddap/"
d_id       <- "EB_MM_BC"
dir_data   <- here("data")
d_csv      <- glue("{dir_data}/{d_id}.csv")
segs_geo   <- glue("{dir_data}/{d_id}_segments.geojson")
d_spp_csv  <- glue("{dir_data}/{d_id}_species.csv")

Sys.setenv(RERDDAP_DEFAULT_URL = erddap_url)
d_info <- info(d_id)

if (!file.exists(d_csv)){
  d <- tabledap(d_info)
  write_csv(d, d_csv)
}
d <- read_csv(d_csv, guess_max = 10000)
```

Of the `r format(nrow(d), big.mark=",") ` rows in this `EB_MM_BC` dataset, let's look at the first 1,000 records:

```{r}
d %>% 
  slice(1:100) %>% 
  datatable()
```

## spatial

### intersect sanctuary with segments

```{r}
xy2ln <- function(x1, y1, x2, y2){
  if (any(is.na(c(x1, y1, x2, y2)))) return(NA)
  
  st_linestring(c(
    st_point(c(x1, y1)), 
    st_point(c(x2, y2))))
}

if (!file.exists(segs_geo)){
  
  segs <- d %>% 
    group_by(
      beach_segment_code, 
      longitude, latitude, 
      north_longitude, north_latitude,
      south_longitude, south_latitude) %>% 
    summarize(
      n_rows = n()) %>% 
    ungroup() %>% 
    mutate(
      geom = pmap(list(
        north_longitude, north_latitude,
        south_longitude, south_latitude), xy2ln) %>% 
        st_sfc(crs = 4326)) %>% 
    st_set_geometry(.$geom)
  
  write_sf(segs, segs_geo, delete_dsn = T)
}
segs <- read_sf(segs_geo)

# TODO: move used fxns into library and utility.R
source("~/github/cinms/scripts/rocky.R")

nms     <- "mbnms"
nms_ply <- get_nms_ply(nms)

# mapview(segs, color="red") + 
#   mapview(nms_ply, alpha.regions = 0.3)

nms_segs <- segs %>% 
  filter(
    st_intersects(
      segs, nms_ply, sparse = F))

m1 <- mapview(nms_ply) + 
  mapview(segs, color="red") + 
  mapview(nms_segs, color="green")
m1
```

NOTE the red segments near San Jose that did not get intersected by the polygon because of spatial coarseness and mismatch.

### intersect buffered sanctuary with segments

Buffer sanctuary by `0.005` decimal degress (~ `r 111 * 0.005` km at equator).

```{r}
nms_ply_buf <- st_buffer(nms_ply, dist = 0.005)

nms_buf_segs <- segs %>% 
  filter(
    st_intersects(
      segs, nms_ply_buf, sparse = F))

m2 <- mapview(nms_ply_buf) +
  mapview(nms_buf_segs, color="green")

sync(m1, m2)
```

## taxa

```{r}
if (!file.exists(d_spp_csv)){
  
  flds_spp <- d_info$variables$variable_name %>% 
    str_subset("species") %>% 
    c("bird_or_mammal")
  
  d_spp <- tabledap(d_info, fields = flds_spp, distinct = T) %>% 
    arrange(bird_or_mammal, species_code)
  
  write_csv(d_spp, d_spp_csv)
}
d_spp <- read_csv(d_spp_csv)

table(d_spp$bird_or_mammal)
```

### birds

```{r}
d_spp %>% 
  filter(bird_or_mammal == "bird") %>% 
  datatable()
```

### mammals

```{r}
d_spp %>% 
  filter(bird_or_mammal == "mammal") %>% 
  datatable()
```


